Study sample characteristics in chronobiology and sleep research

### 2021-05-08 article.Rmd
Updates

---
title             : "Study sample characteristics in chronobiology and sleep research"
shorttitle        : "Study sample characteristics in chronobiology and sleep research"

@@

The extent to which a scientific field's findings are generalisable depend very much on the representativeness of a given study sample. As an example, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020]. The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014].

While research findings converge on participant-level demographic characteristics affecting outcomes, it is not clear to what extent this information is available in the large body of already published literature, nor to what extent it is even reported. Here, we address the question of participant-level demographic characteristics (age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population) and reporting thereof in chronobiology and sleep research. Here, we extracted the study sample characteristics in a total of 1356 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.

@@

*Procedure*. Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered. The listing of possible of target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with >25 years of experience in the field [@spitschan_transparency_2020]. From this previously derived list, we selected eight journals were selected based on their five-year Impact Factor, and included *Journal of Pineal Research* (ISSN: 0742-3098 / 1600-079X; 2018 5-year IF: 12.197), *Sleep*  (0161-8105 / 1550-9109; 5.588), *Journal of Sleep Research* (0962-1105 / 1365-2869; 3.951), *Sleep Medicine* (1389-9457 / 1878-5506; 3.934), *Journal of Clinical Sleep Medicine* (1550-9389 / 1550-9397; 3.855), *Journal of Biological Rhythms* (0748-7304 / 1552-4531; 3.349), Behavioral Sleep Medicine (1540-2002 / 1540-2010; 3.162), and *Chronobiology International* (0742-0528 / 1525-6073; 2.998). While *Sleep Medicine Reviews* also features in the list of journals, we did not include it as it primarily publishes reviews.

*Article inclusion*. 6,777 articles were identified through a MEDLINE search by the journal and including odd years. A random sample of 20% was initially selected for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, errata and corrigenda were excluded.

*Review and article extraction*. All included articles were reviewed for eligibility and coded by RW. The reporting of sample size, age, sex, gender, ethnicity, level of education, socioeconomic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.

@@

*Pre-registration*. We pre-registered our protocol (specified using the PRISMA-P template [@shamseer_preferred_2015; @prisma-p_group_preferred_2015]) on the Open Science Framework (https://osf.io/cu3we/).

*Materials, data and code availability*. All data underlying this manuscript are available on a public GitHub repository (https://github.com/hcvnl/sleep_circadian_demographics_data). The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggExtra; @R-ggformula; @R-ggplot2; @R-ggridges; @R-ggsci; @R-ggstance; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lattice; @R-latticeExtra; @R-lemon; @R-Matrix; @R-mosaic; @R-mosaicData; @R-pracma; @R-readr; @R-readxl; @R-reshape2; @R-stringr; @R-tibble; @R-tidyr; @R-BayesFactor] and is fully reproducible.

@@

# Stacked bar chart of reporting of funding and funding number by year of publication
figure3 <- ggplot(data.funding, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(funding, levels=c("Funding not reported",
                                            "Funding reported",
                                            "Funding and grant reported")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  labs(fill="") +
  scale_fill_manual(values=c("#482677FF","#DCE319FF","#FDE725FF")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

@@

# Stacked bar chart of 8 most represented study countries by year of publication
figure4 <- ggplot(data.country, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(study_country_values, 
                           levels=c("Other","AUS","CAN","DEU","FRA",
                                    "GBR","ITA","JPN","USA")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  labs(fill="Country") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  scale_fill_viridis_d(direction=-1) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10))
```

@@

figure5 <- ggplot(data.ss, aes(x=article_year, y=sample_size_total_reported_values)) +
  geom_violin(fill="#440154FF") +
  theme_apa() +
  scale_y_continuous(trans='log10', breaks=c(10,1000,100000), 
                     labels=c("10","1,000","100,000")) +
  scale_x_discrete(breaks=c("1979","1989","1999","2009","2019")) +
  xlab("Year of publication") +
  ylab("Sample size, log-10 scaling") +
  annotation_logticks(sides="l") +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

@@

# Heat map of reporting of age variables by year of publication
figure6 <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Year of publication") +
  xlab("Age variable") +
  scale_y_continuous(breaks=c(1979,1989,1999,2009,2019), expand=c(0.02,0),
                     labels=c("                   1979", "1989","1999","2009","2019")) +
  scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
                       limits=c(-1,1), labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8), 
        legend.position="right", legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_text(face="bold"))

@@

# Heat map of reporting of age variables by journal of publication
figure7 <- ggplot(data=age.corr.journal, 
                   aes(x=variable, y=article_journal, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Journal of publication") +
  xlab("Age variable") +
  scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
                       limits=c(-1,1), labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.position="right", legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_text(face="bold"))

@@

# Violin plot with regression line and equation for mean age by year of publication
figure8 <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
  geom_smooth(method="lm", color="#440154FF") +
  geom_point(size=1, color="#33638DFF") +
  theme_apa() +
  ylab("Mean age (years)") +
  xlab("Year of publication") +
  scale_x_continuous(limits=c(1979,2019),breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(limits=c(0,80), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

@@

# Stacked bar chart of sex reporting by year of publication
figure9 <- ggplot(data.sex, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(report, levels=c("Unspecified",
                                            "Female",
                                            "Both Sexes",
                                            "Male")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  labs(fill="Sex") +
  scale_fill_manual(values=c("#482677FF","#D8DE29FF","#29AF7FFF","#2D708EFF")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold"),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

@@

figure10 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
  geom_point(size=2.5) +
  geom_line(linetype="dashed") +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  scale_color_viridis_d(name="Variable reported", 
                        labels=c("Education","Ethnicity","Profession","Socio-economic status")) +
  scale_x_continuous(limits=c(1979,2019), breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), limits=c(0,100)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

@@

# Stacked bar chart of the use of demographic variables in the analysis by year of publication
figure12 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
  geom_col(aes(fill=report)) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  labs(fill="Variable reported") +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
```

@@

### Inclusion & Exclusion

  `r round(percent.included)`% of the sampled articles met eligibility criteria during the review.

```{r, fig.cap="Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
figure1
```

```{r, fig.cap="Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
figure2
```

@@

  Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases. Overall, funding by the United States' National Institutes of Health (NIH) represented `r round(percent.funding.NI)`% of the reported funding agencies. `r round(percent.funding.NI.num)`% of the studies funded by the NIH also reported funding number. The second most represented funding agencies were the Australian National Health and Medical Research Council (NHMRC) and the Canadian Institutes of Health Research (CIHR).

```{r, fig.cap="Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
figure3
```

@@

  `r round(100-percent.country.multiple)`% of articles were conducted in a single country. The geographical location of the study was explicitly reported in `r round(percent.country.reported)`% of studies. The country of study was inferred for the rest of the sampled articles. Inference was primarily based on the first author's affiliation. Overall, `r nrow(country.single)` countries were represented. `r nrow(country.multiple)`% of articles reported multiple countries of study. Figure 2B shows the distribution of study location across time with the eight most represented countries.

```{r, fig.cap="Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
figure4
```

@@

```{r, fig.cap="Sample size of the recruited volunteers. Numbers are computed with a log-10 scale.", warning=FALSE, message=FALSE}
figure5
```

@@

  `r round(percent.age)`% of articles reported a variable describing age. Figure 4A and 4B show the use of various age variables by year and journal of publication. Overall, the mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old.

```{r, fig.cap="To be completed.", warning=FALSE, message=FALSE}
figure6
```
```{r, fig.cap="To be completed.", warning=FALSE, message=FALSE}
figure7
```

```{r, fig.cap="Reporting of age variables. Correlation between the reporting of the mean, standrad deviation of the mean, median, minimum, maximum, interquartile range and other age variables by A. year of publication and B. journal of publication. C. Mean age by year of publication.", message=FALSE, warning=FALSE}
figure8
```

@@

  Sex was reported in `r round(percent.sex)`% of the studies. Figure 5 displays the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants. `r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females. Out of the studies focusing on a single gender, `r round(percent.sex.male.focus)`% of the male studies focused on a sex dependent feature, while `r round(percent.sex.female.focus)`% of the female studies did. `r round(percent.sex.age)`% of studies reported age by sex.

```{r, fig.cap="Sex inclusion. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
figure9
```

@@

  Other demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status. Figure 6 shows the distribution of this reporting across the years. Figures 7 show the number of categories reported for each variable.

@@

```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2)
```

@@

  A

@@

@@

>>>>

---
title             : "Inclusion, reporting and analysis of demographic variables in chronobiology and sleep research"
shorttitle        : "Study sample characteristics in chronobiology and sleep research"

@@

The extent to which a scientific field's findings are generalisable depend very much on the representativeness of a given study sample. A recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020]. The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014]. The term "gender data gap" has recently been introduced, demonstrating that women have historically been excluded from biomedical research [@criado-perez_invisible_2020].

The lack of diversity in biomedical and clinical research and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].

While research findings converge on participant-level demographic characteristics affecting outcomes, it is not clear to what extent this information is available in the large body of already published literature, nor to what extent it is even reported. Here, we address the question of participant-level demographic characteristics (age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population) and reporting thereof in chronobiology and sleep research. Here, we extracted the study sample characteristics in a total of 1355 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.

@@

### Procedure

Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered. The listing of possible of target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with >25 years of experience in the field [@spitschan_transparency_2020]. From this previously derived list, we selected eight journals were selected based on their five-year Impact Factor, and included *Journal of Pineal Research* (ISSN: 0742-3098 / 1600-079X; 2018 5-year IF: 12.197), *Sleep*  (0161-8105 / 1550-9109; 5.588), *Journal of Sleep Research* (0962-1105 / 1365-2869; 3.951), *Sleep Medicine* (1389-9457 / 1878-5506; 3.934), *Journal of Clinical Sleep Medicine* (1550-9389 / 1550-9397; 3.855), *Journal of Biological Rhythms* (0748-7304 / 1552-4531; 3.349), Behavioral Sleep Medicine (1540-2002 / 1540-2010; 3.162), and *Chronobiology International* (0742-0528 / 1525-6073; 2.998). While *Sleep Medicine Reviews* also features in the list of journals, we did not include it as it primarily publishes reviews.

### Article inclusion

6,777 articles were identified through a MEDLINE search by the journal and including odd years. A random sample of 20% was initially selected for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, errata and corrigenda were excluded.

### Review and article extraction

All included articles were reviewed for eligibility and coded by RW. The reporting of sample size, age, sex, gender, ethnicity, level of education, socioeconomic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.

@@

### Pre-registration

We pre-registered our protocol (specified using the PRISMA-P template [@shamseer_preferred_2015; @prisma-p_group_preferred_2015]) on the Open Science Framework (https://osf.io/cu3we/).

### Materials, data and code availability

All data underlying this manuscript are available on a public GitHub repository (https://github.com/hcvnl/sleep_circadian_demographics_data). The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggExtra; @R-ggformula; @R-ggplot2; @R-ggridges; @R-ggsci; @R-ggstance; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lattice; @R-latticeExtra; @R-lemon; @R-Matrix; @R-mosaic; @R-mosaicData; @R-pracma; @R-readr; @R-readxl; @R-reshape2; @R-stringr; @R-tibble; @R-tidyr] and is fully reproducible.

@@

# Stacked bar chart of reporting of funding and funding number by year of publication
figure3 <- ggplot(data.funding, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(funding, levels=c("Funding not reported",
                                            "Funding reported",
                                            "Funding and grant reported")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  ggtitle("Funding reporting") +
  labs(fill="") +
  scale_fill_manual(values=c("#482677FF","#DCE319FF","#FDE725FF")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

@@

# Stacked bar chart of 8 most represented study countries by year of publication
figure4 <- ggplot(data.country, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(study_country_values, 
                           levels=c("Other","AUS","CAN","DEU","FRA",
                                    "GBR","ITA","JPN","USA")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  ggtitle("Geographical location of included articles") +
  labs(fill="Country") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  scale_fill_viridis_d(direction=-1) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10))
```

@@

figure5 <- ggplot(data.ss, aes(x=article_year, y=sample_size_total_reported_values)) +
  geom_violin(fill="#440154FF") +
  theme_apa() +
  scale_y_continuous(trans='log10', breaks=c(10,1000,100000), 
                     labels=c("10","1,000","100,000")) +
  scale_x_discrete(breaks=c("1979","1989","1999","2009","2019")) +
  xlab("Year of publication") +
  ylab("Sample size, log-10 scaling") +
  ggtitle("Evolution of sample size") +
  annotation_logticks(sides="l") +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

@@

# Heat map of reporting of age variables by year of publication
figure6 <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Year of publication") +
  xlab("Age variable") +
  ggtitle("Age reporting by year") +
  scale_y_continuous(breaks=c(1979,1989,1999,2009,2019), expand=c(0.02,0),
                     labels=c("                   1979", "1989","1999","2009","2019")) +
  scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
                       limits=c(-1,1), labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8), 
        legend.position="right", legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_text(face="bold"))

@@

# Heat map of reporting of age variables by journal of publication
figure7 <- ggplot(data=age.corr.journal, 
                   aes(x=variable, y=article_journal, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Journal of publication") +
  xlab("Age variable") +
  ggtitle("Age reporting by journal") +
  scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
                       limits=c(-1,1), labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.position="right", legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_text(face="bold"))

@@

# Violin plot with regression line and equation for mean age by year of publication
figure8 <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
  geom_smooth(method="lm", color="#440154FF") +
  geom_point(size=1, color="#33638DFF") +
  theme_apa() +
  ylab("Mean age (years)") +
  xlab("Year of publication") +
  ggtitle("Evolution of mean age by year") +
  scale_x_continuous(limits=c(1979,2019),breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(limits=c(0,80), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

@@

# Stacked bar chart of sex reporting by year of publication
figure9 <- ggplot(data.sex, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(report, levels=c("Unspecified",
                                            "Female",
                                            "Both Sexes",
                                            "Male")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  ggtitle("Sex inclusion by year") +
  labs(fill="Sex") +
  scale_fill_manual(values=c("#482677FF","#D8DE29FF","#29AF7FFF","#2D708EFF")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold"),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

@@

figure10 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
  geom_point(size=2.5) +
  geom_line(linetype="dashed") +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  ggtitle("Reporting of other demographiic variables") +
  scale_color_viridis_d(name="Variable reported", 
                        labels=c("Education","Ethnicity","Profession","Socio-economic status")) +
  scale_x_continuous(limits=c(1979,2019), breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), limits=c(0,100)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

@@

# Stacked bar chart of the use of demographic variables in the analysis by year of publication
figure12 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
  geom_col(aes(fill=report)) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Proportion of articles") +
  ggtitle("Data analysis by demographic variables") +
  labs(fill="Variable reported") +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
```

@@

### Number of analysed articles

Out of `r as.numeric(nrow(data.original))` articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%). The distribution of years in which the articles were published is non-uniform and we included and extracted data from more recent articles (Fig. 1). In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will have not have been available from the entire data collection period (1979 and onwards). We also examined reasons for exclusion amongst the articles that we did not include and extract data from. These are given in Figure 2.

```{r, fig.cap="Included and analysed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
figure1
```

```{r, fig.cap="Normalised distributions of included and excluded articles.", warning=FALSE, message=FALSE}
figure2
```

@@

  Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases (Fig. 3). Overall, funding by the United States' National Institutes of Health (NIH) represented `r round(percent.funding.NI)`% of the reported funding agencies. `r round(percent.funding.NI.num)`% of the studies funded by the NIH also reported funding number. The second most represented funding agencies were the Australian National Health and Medical Research Council (NHMRC) and the Canadian Institutes of Health Research (CIHR).

```{r, fig.cap="Reporting of funding across years.", warning=FALSE, message=FALSE}
figure3
```

@@

  `r round(100-percent.country.multiple)`% of articles were conducted in a single country. The geographical location of the study was explicitly reported in `r round(percent.country.reported)`% of studies. The country of study was inferred for the rest of the sampled articles. Inference was primarily based on the first author's affiliation. Overall, `r nrow(country.single)` countries were represented. `r nrow(country.multiple)`% of articles reported multiple countries of study. Figure 4 shows the distribution of study location across time with the eight most represented countries.

```{r, fig.cap="Gegraphical location of the studies. The eight most represented countries across the entire dataset are individually shown.", warning=FALSE, message=FALSE}
figure4
```

@@

```{r, fig.cap="Sample size of the recruited volunteers as a function of publication year. Numbers are computed on a log-10 scale.", warning=FALSE, message=FALSE}
figure5
```

@@

  `r round(percent.age)`% of articles reported a variable describing age. Overall, the average mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old. We examined the extent to which the mean age across studies differed widely as a function of publication year (Fig. 6), and found that the mean age is much more widely varied in later years, likely reflecting the extent of considering study samples that are more varied in age.

```{r, fig.cap="Evolution of mean age in included studies as a function of publication year. Fit shown is a linear fit.", warning=FALSE, message=FALSE}
figure8
```

@@

  Sex was reported in `r round(percent.sex)`% of the studies. In Figure 7, we show the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants. `r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females. Out of the studies focusing on a single gender, `r round(percent.sex.male.focus)`% of the male studies focused on a sex dependent feature, while `r round(percent.sex.female.focus)`% of the female studies did. `r round(percent.sex.age)`% of studies reported age by sex.

```{r, fig.cap="Sex inclusion by year. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
figure9
```

@@

  We also examined the reporting of other demographic variables, including ethnicity, education, profession and socio-economic status. Other demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status. Figure 8 shows the distribution of this reporting across the years. Qualitatively, we see an increase of reporting additional demographic variables over time. In Figure 9, we show the number of categories reported for each variable amongst those articles that included it in a histogram.

@@

```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2, top="Number of reported categories")
```

@@

  We also examined the extent to which articles reported subgroup analyses of the data based on one or more of the reported demographic variables. Over time (Fig. 10), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed. The most common subgroup analyses involve disaggregting by sex and age, or both.

@@

## Taking an inventory of represented study samples reveals the representativeness of our collective knowledge

The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories. The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020]. In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at these question of participant demographics in chronobiology and sleep research.

##

## The need to consider individual differences

@@

## Limitations of the current review

We now turn to possible limitations of this review and the included analyses and discuss how they might introduce bias in our findings. First, we consider the possibility that the article selection procedure may have introduced biases. Our review only concerned articles from a subset of eight specialized journals. As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialised journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals. This raises the question to what extent we may have missed a section of the literature that would have been relevant to include here. As an alternative strategy, we considered randomly sampling a subset of chronobiology and sleep research articles produced by a general search (e.g. on search from "sleep OR chronobiology" on MEDLINE), but considered this to be too permissive. Our strategy of selecting a subset of candidate journals provided a reasonable trade-off, as well as sampling from a range of field-specific outlets.

Due to the non-uniform distribution of publication years of the included articles (Fig. 1), variables derived from published papers and visualized and/or analyzed by year will have varying uncertainty, with reported percentages from publications of the earlier years being most uncertain. The fact that early years are represented with fewer articles, however, is a not a function of our data set, but of the exponential groth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].

## Towards standardised reporting of demographic variables: From checklists to schemas?

There are guidelines and/or checklists for standardizing reporting of participant characteristics, such as CONSORT [@schulz_consort_2010] or STROBE [@von_elm_strengthening_2007] (an extensive data base for health research reporting guidelines is provided by the Equator Network, https://www.equator-network.org/). Some biomedical journals [e.g. @robinson_transparent_2017] specifically state demographic reporting requirements in the author instructions. Similarly, some organisations may make recommendations of specific reporting items for specific types of study [@veitch_cie_2020].

These guidelines and/or checklists are largely focused on *what* is reported and not *how* it is reported. There is, *a priori*, however, no reason to not develop and use a standardized and machine-readable schema for reporting participant characteristics. The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realising these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics. Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extend such any data schema may be exclusionary (e.g. by enforcing gender binaries), and whether any specific demographic variable is truly important (following the principle of data minimization).
